Think deep for reef resilience

### 2025-07-07 paper/paper_draft.qmd
Reduce use of certain language and specify where coral cover is relative
Specified in figure captions that coral cover is relative to reef area, rather than relative to other reefs or initial conditions. Additionally updated the text on cover timeseries and make clear reminders of where text/citations need to be added.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Current modelling project severe declines in coral cover under projections of Degree Heating Weeks [https://www.biorxiv.org/content/10.1101/2025.01.23.634487v1]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

Individual coral resilience also depends on multiple biological factors beyond connectivity. Coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and inherited thermal tolerance [@humanes2022; @lachs2024] all influence the outcome of high coral heat stress events. Importantly, corals have demonstrated adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios [@bairos-novak2021; @humanes2022]. This adaptive potential represents a crucial mechanism for reef persistence under moderate climate change, working in concert with the environmental factors that modulate thermal stress exposure.

@@

Incorporating the latest advancements in whole ecosystem larval connectivity modeling and coral thermal response and adaptation research into coral population modeling allows an investigation into how abiotic and biotic factors interact to influence coral climate change resilience and the maintenance of positive carbonate budgets. Improvements in coral reef climate response modeling, including sampling from a range of possible climate futures, enable timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015]. While it has previously been suggested that highly connected source reefs that are resilient to bleaching and CoTS outbreaks are likely to play a critical role in supporting future reef recovery across the entire GBR [@hock2017], integrated modelling under current projections of Degree Heating Weeks indicate that depth, rather than connectivity, plays the critical role in determining reef resilience. It is further demonstrated that depth ameliorates reef cover declines and supports maintenance of a postive carbonate budget as the Equilibrium Climate Sensitivity (ECS) of each Global Climate Model (GCM) increases.

@@

ADRIA-CoralBlox was run for 4096 *sampled* scenarios per GCM. A median absolute coral cover timeseries for each reef was calculated for each GCM. For timeseries clustering reef cover timeseries were scaled by the total area of the reef to represent timeseries as a percentage of the total reef area. The proportion of each functional group occupying the reef (group evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declined condition of reefs suffering from very low absolute coral cover.

@@

Each GCM shows similar relationships between reefs in clusters of ‘low’, ‘medium’ and ‘high’ coral cover and contextual analysis factors, thus visual further comparisons of these clusters for each analysis factor are only shown for ACCESS-ESM1-5 GCM (intermediate ECS scenario). Results for the additional GCMs are displayed in the supplementary material. However, there are substantial differences in the longevity of reef resilience and maintenance of coral cover post 2030 between the GCMs, thus coral cover timeseries are displayed for each GCM (displayed at management-area scale to preserve coarse spatial differences in coral trajectories) (@fig-gcm-cover-trajectories).

@@

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2074 under this possible climate future. While recovery phases are predominantly seen in the ‘high’ cover cluster, the ‘medium’ cluster, and to a lesser degree the ‘low’ cluster, also exhibit periods of increasing relative coral cover up to 2045. Under NorESM2-MM GCM the Far Northern management area does not see relative coral cover maintenance as high as the three remaining management regions. Under the second lowest ECS projection (GFDL-CM4, 2.9°C) all clusters in each management area exhibit largely declining future trajectories with periods of recovery from 2030 to 2040. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which results in intermediate levels of coral cover maintenance (between those of lower and higher ECS GCMs). In all management areas reefs in the high cluster are able to maintain relative coral cover levels to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.72°C) and EC-Earth3-Veg (4.31°C) GCMs relative cover trajectories are largely declining, with the exception of ‘high/medium’ clusters in the Cairns-Cooktown management area under EC-Earth3-Veg that display prolonged relative cover maintenance.

![Reef relative coral cover timeseries clustered within each management area, for each GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is represented five times (for each GCM), with each GCM displayed horizontally.](/../outputs/ADRIA_results/HighResCoralStress/figs/GCM_cover_timeseries_plot.png){#fig-gcm-cover-trajectories width=`{julia} timeseries_width`}

@@

Within bioregions timeseries clustering analysis was able to identify three clusters of relative cover timeseries. Across many bioregions timeseries clustering identified a ‘low’ cluster containing reefs that suffer larger declines and generally have lower levels of relative coral cover than the other clusters, a ‘medium’ cluster with intermediate levels of relative coral cover, and a ‘high’ cluster with cover recovery that persists to 2050 in many bioregions (@sfig-access-esm1-5-bioregion-timeseries; @sfig-access-cm2-bioregion-timeseries; @sfig-ec-earth3-veg-bioregion-timeseries; @sfig-gfdl-cm4-bioregion-timeseries; @sfig-noresm2-mm-bioregion-timeseries).

In many bioregions, median reef depth decreases as cluster relative coral cover decreases, with reefs within the ‘high’ cluster commonly having the greatest median reef depths within the bioregion (@fig-access-esm1-5-bioregion-depth; @sfig-access-cm2-bioregion-depth; @sfig-ec-earth3-veg-bioregion-depth; @sfig-gfdl-cm4-bioregion-depth; @sfig-noresm2-mm-bioregion-depth).

![Median depth for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median depth for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/depth_bioregion_violin.png){#fig-access-esm1-5-bioregion-depth width=`{julia} violin_width`}

Correlation between reef total coral cover and reef DHW timeseries decreases as cluster relative coral cover decreases, with reefs in the ‘high’ cluster commonly having the greatest total cover – DHW timeseries correlation levels (@sfig-access-esm1-5-bioregion-dhwcor; @sfig-access-cm2-bioregion-dhwcor; @sfig-ec-earth3-veg-bioregion-dhwcor; @sfig-gfdl-cm4-bioregion-dhwcor; @sfig-noresm2-mm-bioregion-dhwcor).

Conversely, there are no consistent patterns across bioregions in reef relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean; @sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean; @sfig-noresm2-mm-bioregion-dhwmean).
<!--
In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially lower mean thermal tolerance levels compared to reefs in other clusters. However, many bioregions do not show strong differences in thermal tolerance levels (@sfig-access-esm1-5-bioregion-dhwtol; @sfig-access-cm2-bioregion-dhwtol; @sfig-ec-earth3-veg-bioregion-dhwtol; @sfig-gfdl-cm4-bioregion-dhwtol; @sfig-noresm2-mm-bioregion-dhwtol). -->

In many bioregions reefs with high relative coral cover levels (‘high’ cluster) have the lowest initial proportions of habitable area occupied by corals, while reefs in the ‘low’ cluster often have the highest initial proportions of coral habitable area occupied by corals (@sfig-access-esm1-5-bioregion-initialproportion; @sfig-access-cm2-bioregion-initialproportion; @sfig-ec-earth3-veg-bioregion-initialproportion; @sfig-gfdl-cm4-bioregion-initialproportion; @sfig-noresm2-mm-bioregion-initialproportion).

@@

Under all GCMs most reefs suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. Differences between low, medium and high reef clusters in cover timeseries are most notable under these two GCMs.

@@

Overall, our findings do not suggest that reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth and initial coral cover.

@@

Five different GCMs were included in analyses to capture uncertainty in climate system responses to increasing atmospheric carbon dioxide levels. The large differences in relative cover timeseries seen across the GCMs suggest that it is important to consider this uncertainty in analyses of biological climate change responses. This is supported by recent literature suggesting an effective approach to capturing climate change uncertainty is to model climate change responses using a range of climate models separately [@shepherd2018]. However, the patterns identified between low, medium and high clusters of reefs suggest that the importance of depth refugia is consistent across scenarios. Similarly, at the management area scale, only 1% of reefs were assigned a ‘low’ and ‘high’ cluster in different GCMs, suggesting that intrinsic reef properties can prevent high cover resilient reefs from crashing under a different future climate scenario.

@@

![Reef Degree Heating Week timeseries clustered within each management area, for each GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is represented five times (for each GCM), with each GCM displayed horizontally.](/../outputs/ADRIA_results/HighResCoralStress/figs/GCM_dhw_timeseries_plot.png){#sfig-gcm-dhw-trajectories width=`{julia} timeseries_width`}

@@

![Reef relative coral cover timeseries clustered within each bioregion. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/bioregion_cover_timeseries.png){#sfig-access-esm1-5-bioregion-timeseries width=`{julia} timeseries_width`}

![Total coral cover – DHW Pearson’s correlation levels for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median cover – DHW correlation level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_cover_cor_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwcor width=`{julia} violin_width`}

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean width=`{julia} violin_width`}
<!--
![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width=`{julia} violin_width`} -->

![Initial proportion of coral habitable area occupied by corals for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median proportion of habitable area occupied for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/initial_proportion_bioregion_violin.png){#sfig-access-esm1-5-bioregion-initialproportion width=`{julia} violin_width`}

![Log (natural) source – to – sink connectivity ratios for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) source – to – sink ratio for each cluster and grey points represent the ratio value for each reef. Vertical grey lines represent a log (natural) source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/so_to_si_bioregion_violin.png){#sfig-access-esm1-5-bioregion-logsosi width=`{julia} violin_width`}

![Log (natural) total connectivity strength for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/log_total_strength_bioregion_violin.png){#sfig-access-esm1-5-bioregion-totalconnectivity width=`{julia} violin_width`}

@@

![Reef relative coral cover timeseries clustered within each bioregion. ‘Low’ clusters (green) are clusters with the lowest median relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-CM2 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-CM2/bioregion/bioregion_cover_timeseries.png){#sfig-access-cm2-bioregion-timeseries width=`{julia} timeseries_width`}

>>>>

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Current modelling projects severe declines in coral cover under projections of Degree Heating Weeks [**Bozec 2025**]. These coral cover declines will be compounded by ocean acidification, intensifying tropical cyclones, and crown-of-thorns starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs [@cantin2024].

@@

Individual coral resilience depends on multiple biological factors beyond connectivity. Coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and inherited thermal tolerance [@humanes2022; @lachs2024] all influence the outcome of high coral heat stress events. Importantly, corals have demonstrated adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios [@bairos-novak2021; @humanes2022]. This adaptive potential represents a crucial mechanism for reef persistence under moderate climate change, working in concert with the environmental factors that modulate thermal stress exposure.

@@

Incorporating the latest advancements in whole ecosystem larval connectivity modeling and coral thermal response and adaptation research into coral population modeling allows an investigation into how abiotic and biotic factors interact to influence coral climate change resilience and the maintenance of positive carbonate budgets. Improvements in coral reef climate response modeling, including sampling from a range of possible climate futures, enable timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015]. While it has previously been suggested that highly connected source reefs that are resilient to bleaching and CoTS outbreaks are likely to play a critical role in supporting future reef recovery across the entire GBR [@hock2017], integrated modelling under current projections of Degree Heating Weeks indicate that depth, rather than connectivity, plays a critical role in determining reef resilience. It is further demonstrated that depth supports the maintenance of a postive carbonate budget under all Global Climate Models.

@@

ADRIA-CoralBlox was run for 4096 **sampled** scenarios per GCM. A median absolute coral cover timeseries for each reef was calculated for each GCM. For timeseries clustering reef cover timeseries were scaled by the total area of the reef to represent timeseries as a percentage of the total reef area. The proportion of each functional group occupying the reef (group evenness) was also investigated as a metric for timeseries analysis, but was found to not sufficiently capture the declined condition of reefs suffering from very low absolute coral cover.

@@

Each GCM shows similar relationships between reefs in clusters of ‘low’, ‘medium’ and ‘high’ coral cover and contextual analysis factors, thus further comparisons of these clusters for each analysis factor are only shown for ACCESS-ESM1-5 (intermediate ECS scenario). Results for the additional GCMs are displayed in the supplementary material. However, there are substantial differences in the longevity of reef resilience and maintenance of coral cover post 2030 between the GCMs, thus coral cover timeseries are displayed for each GCM (displayed clustered at management-area scale to preserve coarse spatial differences in coral trajectories) (@fig-gcm-cover-trajectories).

NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2075 under this possible climate future. While recovery phases are predominantly seen in the ‘high’ cover cluster, the ‘medium’ cluster, and to a lesser degree the ‘low’ cluster, also exhibit periods of recovery and increasing proportions of coral cover up to 2045. Under NorESM2-MM GCM the Far Northern management area does not see coral cover maintenance as high as the three remaining management regions. Under the second lowest ECS projection (GFDL-CM4, 2.9°C) all clusters in each management area exhibit largely declining future trajectories with periods of recovery from 2030 to 2040. ACCESS-ESM1-5 has an intermediate ECS value of 3.87°C which results in intermediate levels of coral cover maintenance (between those of lower and higher ECS GCMs). In Townsville/Whitsunday and Far Northern management areas reefs in the high cluster are able to maintain coral cover levels around 20% to 2050 under ACCESS-ESM1-5. Under ACCESS-CM2 (4.72°C) and EC-Earth3-Veg (4.31°C) GCMs cover trajectories are largely declining in all management areas.

@@

![Reef area-relative coral cover timeseries clustered within each management area, for each GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is represented five times (for each GCM), with each GCM displayed horizontally.](/../outputs/ADRIA_results/HighResCoralStress/figs/GCM_cover_timeseries_plot.png){#fig-gcm-cover-trajectories width=`{julia} timeseries_width`}

@@

Within bioregions timeseries clustering analysis was able to identify three clusters of area-relative cover timeseries. Across many bioregions timeseries clustering identified a ‘low’ cluster containing reefs that suffer larger declines and generally have lower levels of coral cover than the other clusters, a ‘medium’ cluster with intermediate levels of coral cover, and a ‘high’ cluster with greater levels of cover recovery, with persistence to 2050 in some bioregions (@sfig-access-esm1-5-bioregion-timeseries; @sfig-access-cm2-bioregion-timeseries; @sfig-ec-earth3-veg-bioregion-timeseries; @sfig-gfdl-cm4-bioregion-timeseries; @sfig-noresm2-mm-bioregion-timeseries).

In many bioregions, median reef depth decreases as cluster coral cover decreases, with reefs within the ‘high’ cluster commonly having the greatest median reef depths within the bioregion (@fig-access-esm1-5-bioregion-depth; @sfig-access-cm2-bioregion-depth; @sfig-ec-earth3-veg-bioregion-depth; @sfig-gfdl-cm4-bioregion-depth; @sfig-noresm2-mm-bioregion-depth).

![Median depth for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median depth for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/depth_bioregion_violin.png){#fig-access-esm1-5-bioregion-depth width=`{julia} violin_width`}

Correlation between reef total coral cover and reef DHW timeseries decreases as cluster coral cover decreases, with reefs in the ‘high’ cluster commonly having the greatest total cover – DHW timeseries correlation levels (@sfig-access-esm1-5-bioregion-dhwcor; @sfig-access-cm2-bioregion-dhwcor; @sfig-ec-earth3-veg-bioregion-dhwcor; @sfig-gfdl-cm4-bioregion-dhwcor; @sfig-noresm2-mm-bioregion-dhwcor).

Conversely, there are no consistent patterns across bioregions in reef area-relative cover and mean timeseries DHW levels, with clusters having similar mean DHW levels in many bioregions (@sfig-access-esm1-5-bioregion-dhwmean; @sfig-access-cm2-bioregion-dhwmean; @sfig-ec-earth3-veg-bioregion-dhwmean; @sfig-gfdl-cm4-bioregion-dhwmean; @sfig-noresm2-mm-bioregion-dhwmean).
<!-- 
In some bioregions (4 for ACCESS-ESM1-5) reefs in the ‘low’ cluster have substantially lower mean thermal tolerance levels compared to reefs in other clusters. However, many bioregions do not show strong differences in thermal tolerance levels (@sfig-access-esm1-5-bioregion-dhwtol; @sfig-access-cm2-bioregion-dhwtol; @sfig-ec-earth3-veg-bioregion-dhwtol; @sfig-gfdl-cm4-bioregion-dhwtol; @sfig-noresm2-mm-bioregion-dhwtol). -->

In many bioregions reefs with high area-relative coral cover levels (‘high’ cluster) have the lowest initial proportions of habitable area occupied by corals, while reefs in the ‘low’ cluster often have the highest initial proportions of coral habitable area occupied by corals (@sfig-access-esm1-5-bioregion-initialproportion; @sfig-access-cm2-bioregion-initialproportion; @sfig-ec-earth3-veg-bioregion-initialproportion; @sfig-gfdl-cm4-bioregion-initialproportion; @sfig-noresm2-mm-bioregion-initialproportion).

@@

Under all GCMs most reefs are projected to suffer substantial declines in coral cover from 2025 levels. NorESM2-MM and ACCESS-ESM1-5 GCMs convey the most coral cover maintanance into the future. Differences between low, medium and high reef clusters in cover timeseries are most notable under these two GCMs. Maintainance of coral cover into the future is not clearly linked to the Equilibrium Climate Sensitivity of each GCM used as ACCESS-ESM1-5 has a higher ECS value than GFDL-CM4, with reefs experiencing large continuous declines under GFDL-CM4.

@@

Overall, our findings do not suggest that reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters **need to replot tolerance values**. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth.

@@

Five different GCMs were included in analyses to capture uncertainty in climate system responses to increasing atmospheric carbon dioxide levels. The large differences in cover timeseries seen across the GCMs suggest that it is important to consider this uncertainty in analyses of biological climate change responses. This is supported by recent literature suggesting an effective approach to capturing climate change uncertainty is to model climate change responses using a range of climate models separately [@shepherd2018]. However, the patterns identified between low, medium and high clusters of reefs suggest that the importance of depth refugia is consistent across scenarios. Similarly, at the management area scale, only 1% of reefs were assigned a ‘low’ and ‘high’ cluster in different GCMs, suggesting that intrinsic reef properties can prevent high cover resilient reefs from crashing under a different future climate scenario.

@@

![Reef Degree Heating Week timeseries clustered within each management area, for each GCM analysed. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Each management area is represented five times (for each GCM), with each GCM displayed horizontally.](/../outputs/ADRIA_results/HighResCoralStress/figs/GCM_dhw_timeseries_plot.png){#sfig-gcm-dhw-trajectories width=`{julia} timeseries_width`}

@@

![Reef area-relative coral cover timeseries clustered within each bioregion. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/bioregion_cover_timeseries.png){#sfig-access-esm1-5-bioregion-timeseries width=`{julia} timeseries_width`}

![Total coral cover – DHW Pearson’s correlation levels for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median cover – DHW correlation level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_cover_cor_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwcor width=`{julia} violin_width`}

![Mean timeseries DHW values for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW level for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/mean_dhw_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwmean width=`{julia} violin_width`}
<!--
![Mean timeseries DHW tolerance value for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median DHW tolerance value for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/dhw_tolerance_bioregion_violin.png){#sfig-access-esm1-5-bioregion-dhwtol width=`{julia} violin_width`} -->

![Initial proportion of coral habitable area occupied by corals for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the median proportion of habitable area occupied for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/initial_proportion_bioregion_violin.png){#sfig-access-esm1-5-bioregion-initialproportion width=`{julia} violin_width`}

![Log (natural) source – to – sink connectivity ratios for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) source – to – sink ratio for each cluster and grey points represent the ratio value for each reef. Vertical grey lines represent a log (natural) source – to – sink ratio of 0 indicating equal outgoing and incoming connectivity levels. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/so_to_si_bioregion_violin.png){#sfig-access-esm1-5-bioregion-logsosi width=`{julia} violin_width`}

![Log (natural) total connectivity strength for reefs within each bioregion, separated into clusters assigned by timeseries clustering. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines within each violin represent the median log (natural) total connectivity strength for each cluster. Displays results for ACCESS-ESM1-5 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-ESM1-5/bioregion/log_total_strength_bioregion_violin.png){#sfig-access-esm1-5-bioregion-totalconnectivity width=`{julia} violin_width`}

@@

![Reef area-relative coral cover timeseries clustered within each bioregion. ‘Low’ clusters (green) are clusters with the lowest median area-relative cover between 2030 and 2040 within each bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative cover between 2030 and 2040 within each bioregion. ‘High’ clusters (blue) are clusters with the highest median area-relative cover between 2030 and 2040 within each bioregion. Solid lines represent the cluster median timeseries and bands represent the 2.5 to 97.5 percentile confidence intervals. Displays results for ACCESS-CM2 GCM.](/../outputs/ADRIA_results/HighResCoralStress/figs/ACCESS-CM2/bioregion/bioregion_cover_timeseries.png){#sfig-access-cm2-bioregion-timeseries width=`{julia} timeseries_width`}
